Calendar-based graphics for visualising people's daily schedules

### 2017-07-12 main.Rmd
Added opening paragraph to section 3

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may becomes less perceivable of the actual variations or behaviours from the viewer's perspective.

@@

The algorithms to construct calendar grids, several scales, as well as reference lines and labels are described in the section.

@@

We only illustrate that grids are laid out over the horizontal direction in the paper. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above. It particularly benefits those users who get accustomed to calendars of vertical organisation in some countries.

@@

The \code{frame_calendar} function assumes that the underlying data structure is tidy data [@wickham2014tidy].

@@

@@

The algorithm can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as "weekly" and "daily" calendar respectively. Due to the way of the arrangement, the weekly calendar puts more emphases on days of a week over days of a year, whereas the daily one serves as the opposite. The monthly format can be considered as the advanced twist of both weekly and daily. Temporal patterns lead to which format to be employed. The weekly calendar can be a nice attempt if the most variations can be characterised by days of a week. On the other hand, the absence of weekly effect but the presence of yearly effect can direct to the daily calendar. When both are present, the monthly calendar appears to be a better choice.

## Polar coordinates

@@

Grouping the cells according to the common period give rise to these scales. The minimums and maximums obtained for every $(i, j)$ and $(m, n)$ allow for individual scale over all the single days. Similarly, $j$ indexing day of the week is grouped to determine the scales on days of a week, and $g$ representing the day in the month is for the scales on days of a month.

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $m$ and $n$. The weekday texts with a single letter are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$, with the central position of a cell $x / 2$ for each $j$.

@@

The \code{frame_calendar} does not constrain itself to mapping only temporal variable to the \code{x}. Figure \ref{fig:scatterplot} shows the lag scatterplot where hourly counts are assigned to the \code{x} and lagged themselves to the \code{y}. It provides a visual tool for identifying repeating patterns.

>>>>

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 3 sensors. Figure \ref{fig:facet-time}, which is complementary to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

Calendar-based graphics turn out to be a useful tool in unfolding human-related activities in temporal context. For example, @VanWijkCluster1999 developed a calendar view of heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It remarkably contrasts weekdays and weekends, highlights public holiday, and presents other known seasons like school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in a couple of R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are too constrained to colour-encoding graphics and day of week as the smallest time scale. Time of day, which serves as one of the most important aspect in explaining variations arisen from pedestrian sensor data, will be neglected through daily aggregation. Additionally if simply using colours encoded as quantities, it may become less perceivable of the actual variations or behaviours from the viewer's perspective.

@@

@@

We only illustrate that grids are laid out over the horizontal direction in the paper. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above when the argument \code{dir} is set to \code{"v"}. It particularly benefits those users who get accustomed to calendars of vertical organisation in some countries.

@@

There are a few options provided for the \code{frame_calendar} function to adjust the display and they are shown as follows:

@@

Assuming that tidy data [@wickham2014tidy] is the underlining data structure, the \code{x} takes a variable that will be mapped to the x axis and the \code{y} mapped to the y axis for the later plotting. In Figure \ref{fig:flinders-2016}, for example, the \code{x} is the variable specifying the time of the day, and the \code{y} is the variable suggesting the hourly counts. The \code{date} argument is given by the date variable that determines the correct order in the calendar layout. We shall describe some of the arguments that control the graphics in different ways.

@@

The algorithm described in Section \ref{sec:algorithm} is for the most common calendar layout--"monthly" calendar, and it can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as "weekly" and "daily" calendar respectively. Which layout to be used is controlled by the \code{calendar} argument. Due to the way of the arrangement, the weekly calendar puts more emphases on days of a week over days of a year, whereas the daily one serves as the opposite. The monthly format can be considered as the advanced twist of both weekly and daily. Temporal patterns lead to which format to be employed. The weekly calendar can be a nice attempt if the most variations can be characterised by days of a week. On the other hand, the absence of weekly effect but the presence of yearly effect can direct to the daily calendar. When both are present, the monthly calendar appears to be a better choice.

## Polar transformation

When \code{polar = TRUE}, the polar transformation is carried out for the data.

@@

Grouping the cells according to the common period gives rise to these scales. The minimums and maximums obtained for every $(i, j)$ and $(m, n)$ allow for individual scale over all the single days. Similarly, $j$ indexing day of the week is grouped to determine the scales on days of a week, and $g$ representing the day in the month is for the scales on days of a month.

@@

The abbreviated month labels located on the top left are obtained through $(\min{(x)}, \max{(y)})$ for every $(m, n)$. The weekday texts with a single letter are uniformly placed on the bottom of the whole canvas, that is $\min{(y)}$, with the central position of a cell $x / 2$ for each $j$.

@@

The \code{frame_calendar} function does not constrain itself to mapping only temporal variable to the \code{x}. Figure \ref{fig:scatterplot} shows the lag scatterplot where hourly counts are assigned to the \code{x} and lagged themselves to the \code{y}. It provides a visual tool for identifying repeating patterns.

### 2017-07-12 main.tex
Added opening paragraph to section 3

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arising from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using colours encoded as quantities, it may becomes less
perceivable of the actual variations or behaviours from the viewer's
perspective.

@@

We only illustrate that grids are laid out over the horizontal direction
in the paper. The vertical direction can be enabled by swapping \(i\)
and \(j\) in the algorithm stated above. It particularly benefits those
users who get accustomed to calendars of vertical organisation in some
countries.

@@

The \code{frame_calendar} function assumes that the underlying data
structure is tidy data \citep{wickham2014tidy}.

@@

@@

The algorithm can be simplified to accommodate the other two types of
calendar formats. One is comprised of days of a week in columns and
weeks of a year in rows, and the other is days of a month in columns and
months of a year in rows, which we refer to as ``weekly'' and ``daily''
calendar respectively. Due to the way of the arrangement, the weekly
calendar puts more emphases on days of a week over days of a year,
whereas the daily one serves as the opposite. The monthly format can be
considered as the advanced twist of both weekly and daily. Temporal
patterns lead to which format to be employed. The weekly calendar can be
a nice attempt if the most variations can be characterised by days of a
week. On the other hand, the absence of weekly effect but the presence
of yearly effect can direct to the daily calendar. When both are
present, the monthly calendar appears to be a better choice.

\subsection{Polar coordinates}\label{polar-coordinates}

@@

Grouping the cells according to the common period give rise to these
scales. The minimums and maximums obtained for every \((i, j)\) and
\((m, n)\) allow for individual scale over all the single days.
Similarly, \(j\) indexing day of the week is grouped to determine the
scales on days of a week, and \(g\) representing the day in the month is
for the scales on days of a month.

@@

The abbreviated month labels located on the top left are obtained
through \((\min{(x)}, \max{(y)})\) for every \(m\) and \(n\). The
weekday texts with a single letter are uniformly placed on the bottom of
the whole canvas, that is \(\min{(y)}\), with the central position of a
cell \(x / 2\) for each \(j\).

>>>>

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 3 sensors. Figure \ref{fig:facet-time}, which is
complementary to the Figure \ref{fig:time-series-plot}, provides another
aspect of temporal patterns in the more detailed level by faceting day
of the week. It lends itself to a number of issues that make exploratory
data visualisation challenging in many temporal-context applications
involving human behaviours:

@@

Calendar-based graphics turn out to be a useful tool in unfolding
human-related activities in temporal context. For example,
\citet{VanWijkCluster1999} developed a calendar view of heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It remarkably
contrasts weekdays and weekends, highlights public holiday, and presents
other known seasons like school vacations, all of which have influence
over the turn-outs in the office. The calendar-based heatmap was
implemented in a couple of R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are too constrained to colour-encoding graphics and day of
week as the smallest time scale. Time of day, which serves as one of the
most important aspect in explaining variations arisen from pedestrian
sensor data, will be neglected through daily aggregation. Additionally
if simply using colours encoded as quantities, it may become less
perceivable of the actual variations or behaviours from the viewer's
perspective.

@@

We only illustrate that grids are laid out over the horizontal direction
in the paper. The vertical direction can be enabled by swapping \(i\)
and \(j\) in the algorithm stated above when the argument \code{dir} is
set to \code{"v"}. It particularly benefits those users who get
accustomed to calendars of vertical organisation in some countries.

@@

There are a few options provided for the \code{frame_calendar} function
to adjust the display and they are shown as follows:

@@

Assuming that tidy data \citep{wickham2014tidy} is the underlining data
structure, the \code{x} takes a variable that will be mapped to the x
axis and the \code{y} mapped to the y axis for the later plotting. In
Figure \ref{fig:flinders-2016}, for example, the \code{x} is the
variable specifying the time of the day, and the \code{y} is the
variable suggesting the hourly counts. The \code{date} argument is given
by the date variable that determines the correct order in the calendar
layout. We shall describe some of the arguments that control the
graphics in different ways.

@@

The algorithm described in Section \ref{sec:algorithm} is for the most
common calendar layout--``monthly'' calendar, and it can be simplified
to accommodate the other two types of calendar formats. One is comprised
of days of a week in columns and weeks of a year in rows, and the other
is days of a month in columns and months of a year in rows, which we
refer to as ``weekly'' and ``daily'' calendar respectively. Which layout
to be used is controlled by the \code{calendar} argument. Due to the way
of the arrangement, the weekly calendar puts more emphases on days of a
week over days of a year, whereas the daily one serves as the opposite.
The monthly format can be considered as the advanced twist of both
weekly and daily. Temporal patterns lead to which format to be employed.
The weekly calendar can be a nice attempt if the most variations can be
characterised by days of a week. On the other hand, the absence of
weekly effect but the presence of yearly effect can direct to the daily
calendar. When both are present, the monthly calendar appears to be a
better choice.

\subsection{Polar transformation}\label{polar-transformation}

When \code{polar = TRUE}, the polar transformation is carried out for
the data.

@@

Grouping the cells according to the common period gives rise to these
scales. The minimums and maximums obtained for every \((i, j)\) and
\((m, n)\) allow for individual scale over all the single days.
Similarly, \(j\) indexing day of the week is grouped to determine the
scales on days of a week, and \(g\) representing the day in the month is
for the scales on days of a month.

@@

The abbreviated month labels located on the top left are obtained
through \((\min{(x)}, \max{(y)})\) for every \((m, n)\). The weekday
texts with a single letter are uniformly placed on the bottom of the
whole canvas, that is \(\min{(y)}\), with the central position of a cell
\(x / 2\) for each \(j\).
